# Reproducible Research: Peer Assessment 1

Work by mvelascoc (2014/08/10), avaliable in [https://github.com/mvelascoc/RepData_PeerAssessment1](https://github.com/mvelascoc/RepData_PeerAssessment1)

## Loading and preprocessing the data
```{r}
  ## Read de complete file
  file <- "activity.csv"
  data <- read.csv(file, header = TRUE)
  ## ignore incomplete data
  good <- complete.cases(data[,1])
  subdata <- data[good,]
  ## use date as factor to separate the data. This creates a list in which
    ## data from each day is separated
  stepsByDay <- split(subdata, subdata$date)
```


## What is mean total number of steps taken per day?
Data to be processed is the sum of the recorded steps each day. I create a table with this data.
Note that the NA values were taken before calculating values requieres, but the days with 0 values were not. **I do not consider 0 to be a missing value** in this exercise: person could be asleep all day or in bed, I have no information to decide its wrong data. I am only taking away NA values.

```{r}
  ## Calculate the steps recorded per day by adding the steps recorded in each
    ## interval of the day, and save it in one single vector
  totalSteps <- data.frame(names(stepsByDay), 0)
  colnames(totalSteps) <- c("Day", "Total Steps Recorded")
  for (i in 1:length(stepsByDay)) {
      dimDay <- dim(stepsByDay[[i]])
      if (dimDay[1] > 0) {
         totalStepsDay <- sum(stepsByDay[[i]][,1])
         totalSteps[i,2] <- totalStepsDay
      }
  }

```

Data prepared is shown below:

```{r echo = FALSE}
  print(totalSteps)
```
Now, I calculate the descriptive statistics required.
```{r}
## Calculating mean and Median
  meanSteps <- mean(totalSteps[,2])
  medianSteps <- median(totalSteps[,2])
```
The mean total number of steps taken per day is `meanSteps` and median in `medianSteps`. Its values are shown:
```{r}
  print(paste ("Mean (value of meanSteps): ", meanSteps))
  print(paste ("Median (value of medianSteps): ", medianSteps))
  
  hist(totalSteps[,2], main = "Total Steps Recorded By Day", 
       xlab = "Total Steps Recorded", col = "turquoise4", breaks = 10)

```

## What is the average daily activity pattern?

Daily activity required is the average for each interval. I will again use the prepared list `stepsByDay` that contains activity for each day to create a table.

Again, please note that I am not considering the values 0 as missing values. 

```{r}

  ## take steps from one of the samples
  intFac <- factor(stepsByDay[[2]][,3], ordered = TRUE)
  interv <- levels(intFac)
  meanStepsByRange <- data.frame(interv, 0, 0)
  colnames(meanStepsByRange) <- c("Range", 
                                  "Sum Total Steps Recorded", "Average Steps")
  for (i in 1:length(stepsByDay)) {  ##  for all days
      dimDay <- dim(stepsByDay[[i]])
      if (dimDay[1] > 0) {  ## if there is data in this day 
         dimDay[1]
         for (j in 1: dimDay[1]) {  ## for all ranges
            range <-  stepsByDay[[i]][j,3]
            steps <- stepsByDay[[i]][j,1]
            meanStepsByRange[meanStepsByRange$Range == range, 2] <- meanStepsByRange[meanStepsByRange$Range == range, 2] + steps    ## I add the steps
         }
         
      }
  }
  ## when the adding is over, the average for all days is calculated
  meanStepsByRange[,3] <- meanStepsByRange[,2] / length(stepsByDay)


```

The average pattern for the steps is shown in the following graph
```{r}
  ## pattern is now shown in a graph  
  plot(interv, meanStepsByRange[,3], type = "l", 
       main = "Average Steps by interval", xlab = "interval", ylab = "steps")
   
```

The interval in which the average is the maximum is calculated and printed with
```{r}
    maxSteps <- max(meanStepsByRange[,3])
    index <- which(meanStepsByRange[,3] == maxSteps, arr.ind=TRUE)
    text <- paste("Maximum value is ", maxSteps, 
                  " and its found in the interval ", meanStepsByRange[index,1])
   print(text)
```

## Imputing missing values

Since NA values were taken from the initial sample, vector `good` contains values that need to be replaced. I will create a new set of data incluing the mean value for the interval, rounded to next integer, calculated in the last section of the exercise. 

```{r}
    ## calculate and print number of NA values
    goodValues <- as.numeric(!good)
    totalNAs <- sum(goodValues)
    text <- paste ("Number of NA data fields is ", totalNAs)
    print(text)

    ## create a new vector in which NA data is replaced by average in the
    ## same interval
    dataWithNA <- data
    for (k in 1:dim(data)[1]){
        if (!(good[k])){  ## has a NA value
            index <- which(meanStepsByRange[,1] == data[k,3], arr.ind=TRUE)
            dataWithNA[k,1] <- round(meanStepsByRange[index,3])
        }
    }

  ## use date as factor to separate the data. This creates a list in which
    ## data from each day is separated, so processing is easier
  stepsByDayWithNA <- split(dataWithNA, dataWithNA$date)

    ## Now I repeat the same processing to get a histogram 
## Calculate the steps recorded per day by adding the steps recorded in each
    ## interval of the day, and save it in one single vector
  totalSteps2 <- data.frame(names(stepsByDayWithNA), 0)
  colnames(totalSteps2) <- c("Day", "Total Steps Recorded")
  for (i in 1:length(stepsByDayWithNA)) {
      dimDay2 <- dim(stepsByDayWithNA[[i]])
      if (dimDay2[1] > 0) {
         totalStepsDay2 <- sum(stepsByDayWithNA[[i]][,1])
         totalSteps2[i,2] <- totalStepsDay2
      }
  }

  meanSteps2 <- mean(totalSteps2[,2])
  medianSteps2 <- median(totalSteps2[,2])


  par(mfrow = c(1,2))

  hist(totalSteps[,2], main = "Total Steps Recorded By Day", 
       xlab = "Total Steps Recorded", col = "turquoise4", breaks = 10)

  hist(totalSteps2[,2], main = "Total Steps Recorded By Day, with NA replaced", 
       xlab = "Total Steps Recorded", col = "salmon", breaks = 10)


  print(paste ("Mean was: ", meanSteps, " and now its ", meanSteps2))
  print(paste ("Median was: ", medianSteps, " and now its" ,medianSteps2))


```
The value of the mean changes, but the median doesn't. The strategy used to fill the NAs used the mean value of the interval was lower that the mean value. 
This can also be verified in the histogram change.

## Are there differences in activity patterns between weekdays and weekends?

To answer this question I am using the code used to plot the pattern, whith a change to separete weekdays from weekends. 

```{r}

  ## take steps from one of the samples
  intFac2 <- factor(stepsByDay[[2]][,3], ordered = TRUE)
  interv2 <- levels(intFac2)
  meanStepsByRange2 <- data.frame(interv2, 0, 0, 0, 0)
  colnames(meanStepsByRange2) <- c("Range",  "Sum_Steps_Recorded_W", "Average_Steps_W", "Sum_Steps_Recorded_WEND", "Average_Steps_WEND")
  
  weekDaysNum <- 0   # counters of types of days, to be used to evaluate average
  weekendDaysNum <- 0
  Sys.setlocale("LC_TIME", "en_US") #make sure I can compare weekday names

  for (i in 1:length(stepsByDay)) {  ##  for all days
      dimDay <- dim(stepsByDay[[i]])
      if (dimDay[1] > 0) {  ## if there is data in this day 
         
         for (j in 1: dimDay[1]) {  ## for all ranges
             day <- strptime(stepsByDay[[i]][j,2], "%Y-%m-%d")
             dayName <- strftime(day, "%a")
             range <-  stepsByDay[[i]][j,3]
             steps <- stepsByDay[[i]][j,1]
              
             if (is.element( dayName, c("Sun","Sat"))) { #its a day of weekend
                weekendDaysNum <- weekendDaysNum + 1
               meanStepsByRange2[meanStepsByRange2$Range == range, 4] <- meanStepsByRange2[meanStepsByRange2$Range == range, 4] + steps    ## I add the steps
            } else {  # its a weekday
                weekDaysNum <- weekDaysNum + 1
               meanStepsByRange2[meanStepsByRange2$Range == range, 2] <- meanStepsByRange2[meanStepsByRange2$Range == range, 2] + steps    ## I add the steps

            }
            
         }
         
      }
  }
  ## when the adding is over, the average for all days is calculated
  meanStepsByRange2[,5] <- meanStepsByRange2[,4] / weekendDaysNum
  meanStepsByRange2[,3] <- meanStepsByRange2[,2] / weekDaysNum

```

The average pattern for the steps for Weekdays and Weekend days is shown in the following graph
```{r}
  ## pattern is now shown in a graph  
    par(mfrow = c(1,2))
    plot(interv2, meanStepsByRange2[,5], type = "l", 
    main = "Average Steps by interval-Weekend", xlab = "interval", ylab = "steps")
    plot(interv2, meanStepsByRange2[,3], type = "l", 
    main = "Average Steps by interval-Week", xlab = "interval", ylab = "steps")
```

In order to compare them easily, the averages are plotted togheter and a predictor is included

```{r}

    loWEND <- loess(meanStepsByRange2[,5]~interv2)
    loW <- loess(meanStepsByRange2[,3]~interv2)
    
    plot(interv2, meanStepsByRange2[,5], type = "l", col='red',
    main = "Average Steps by Interval", xlab = "interval", ylab = "steps")
    
    lines(interv2, predict(loWEND), col='red', lwd=4, type="l")
    lines(interv2, predict(loW), col='blue', lwd=4, type="l")
    lines(interv2, meanStepsByRange2[,3], type= "l", col="blue")
      
    legend(x="topright", legend=c("Weekend", "Weekdays"), lwd=2,
             col=c("red", "blue"))

```
Both graphics shows that there is a significant difference in the pattern for weekdays and weekends.   
Weekdends include a significant peak around interval 700 but with more activity during the whole day and centered in the 1500 interval. Week days have a peak that starts earlier, is also around 700 interval but has a significantly lower value though out the day.